---
title: "dsm"
author: "Jez Bird"
date: "28 March 2019"
output: html_document
---

# Density Surface Modelling

Pages 1002-1003 of Miller et al. 2013 give an excellent summary of approaches to spatial modelling of distance sampling data. They advocate density surface modelling as a preferred option but it would be good to discuss this (and to potentially adopt alternative approaches for modelling Blue Petrel and Grey Petrel data).

I've had a preliminary but fairly weak effort at running some basic density surface models on our data. I don't really understand the outputs at this point and am stuck at the point of needing to run predictions from the model. 

The transects have been split into roughly 20 m square segments. I've calculate mean elevation and mean slope for each segment. We now have segment-level data, distance-data and linked distance-observations to segments.

```{r}
knitr::opts_knit$set(root.dir = normalizePath("C:/Users/Jez/Documents/PhD/Data/Island-wide survey/R/dsm"))
library(dsm)
library(Distance)
## Loading required package: mgcv
## Loading required package: nlme
## This is mgcv 1.8-7. For overview type 'help("mgcv-package")'.
## Loading required package: mrds
## This is mrds 2.1.15
## Built: R 3.2.2; ; 2015-11-24 17:46:38 UTC; unix
## This is dsm 2.2.11
## Built: R 3.2.2; ; 2015-11-24 18:04:45 UTC; unix
library(ggplot2)
library(knitr)
library(captioner)
library(rgdal)
library(plyr)
library(maptools)
library(tweedie)

# plotting options
gg.opts <- theme(panel.grid.major=element_blank(),
                 panel.grid.minor=element_blank(),
                 panel.background=element_blank())

# make the results reproducible
set.seed(11123)

segdata <- read.csv("segdata.csv")

distdata <- read.csv("distdata.csv")

obsdata <- read.csv("obsdata.csv")

```

We generate a detection function, but this should eventually come out of the previous analysis including covariates.

```{r}

detfc.hr.null<-ds(distdata, max(distdata$distance), key="hr", adjustment=NULL)
```

Then fit a simple dsm

```{r}

dsm.xy <- dsm(count~s(x,y), detfc.hr.null, segdata, obsdata, method="REML")

summary(dsm.xy)

vis.gam(dsm.xy, plot.type="contour", view=c("x","y"), asp=1, type="response", contour.col="black", n.grid=100)
```

Then more complex models with elevation 

```{r}

dsm.xy.elevation <- dsm(count~s(x,y,k=10) + s(elevation_mean,k=20), detfc.hr.null, segdata, obsdata, group = TRUE, segment.area = "area", method="REML")
summary(dsm.xy.elevation)

plot(dsm.xy.elevation, select=2)

```

and elevation and slope

```{r}

dsm.xy.ele.slope <- dsm(count~s(x,y,k=10) + s(elevation_mean,k=20) +s(slope_mean, k=20), detfc.hr.null, segdata, obsdata, group = TRUE, segment.area = "area", method="REML")
summary(dsm.xy.ele.slope)

plot(dsm.xy.ele.slope, select=c(2,3))

gam.check(dsm.xy)
gam.check(dsm.xy.ele.slope)
```

The next stage will be to make predictions from this data.

